\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{titlesec}

\hypersetup{
  colorlinks = true,
  linkcolor  = blue!60!black,
  urlcolor   = blue!60!black,
  citecolor  = blue!60!black,
}

% --- Code listing style for shell commands -----------------------------------
\definecolor{shellbg}{HTML}{F5F5F5}
\definecolor{shellrule}{HTML}{CCCCCC}
\definecolor{shellcomment}{HTML}{6A737D}
\definecolor{shellkeyword}{HTML}{005CC5}

\lstdefinestyle{shell}{
  basicstyle      = \ttfamily\footnotesize,
  backgroundcolor = \color{shellbg},
  frame           = single,
  rulecolor       = \color{shellrule},
  framesep        = 6pt,
  xleftmargin     = 6pt,
  xrightmargin    = 6pt,
  breaklines      = true,
  showstringspaces= false,
  columns         = fullflexible,
  keepspaces      = true,
  commentstyle    = \color{shellcomment}\itshape,
  keywordstyle    = \color{shellkeyword}\bfseries,
  morecomment     = [l]{\#},
  morekeywords    = {sudo, apt, apt-get, yum, dnf, brew, git, make, cd, mkdir,
                     export, source, ./configure, configure, echo, tar, wget,
                     curl, ldconfig, which, ln, cp, rm, mv},
  literate        = {/}{{/}}1 {-}{{-}}1 {=}{{=}}1,
}
\lstset{style=shell}

\titleformat{\section}{\Large\bfseries}{\thesection}{0.7em}{}
\titleformat{\subsection}{\large\bfseries}{\thesubsection}{0.6em}{}

\title{Installing Magic, Netgen, IRSIM, ngspice, xschem, and open\_pdks\\
       \large An installation guide for the Open Circuit Design tool suite}
\author{James E. Stine and Landon Burleson}
\date{\today}

\begin{document}
\maketitle

\section{Overview}

This guide walks through building and installing the open-source analog/digital
IC design flow maintained primarily by Tim Edwards into a self-contained user
prefix. The default location used throughout this document is

\begin{lstlisting}
$HOME/opencircuitdesign
\end{lstlisting}

\noindent which matches the default \texttt{OPEN\_CIRCUITDESIGN\_ROOT} value
in the included \texttt{setup.sh}. If you would rather install elsewhere
(e.g.\ \texttt{/opt/opencircuitdesign} or \texttt{/home/user/ocd}), simply
edit the line near the top of \texttt{setup.sh}:

\begin{lstlisting}
export OPEN_CIRCUITDESIGN_ROOT="$HOME/opencircuitdesign"
\end{lstlisting}

\noindent and use that same path everywhere this guide writes
\texttt{\$OPEN\_CIRCUITDESIGN\_ROOT}.

The tools covered are:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Magic} --- VLSI layout editor (\url{http://opencircuitdesign.com/magic}).
  \item \textbf{Netgen} --- netlist comparison / LVS
        (\url{http://opencircuitdesign.com/netgen}).
  \item \textbf{IRSIM} --- switch-level digital simulator
        (\url{http://opencircuitdesign.com/irsim}).
  \item \textbf{ngspice} --- analog/mixed-signal SPICE simulator
        (\url{https://ngspice.sourceforge.io}). \emph{Note:} ngspice is a
        separate project from Tim Edwards' tools, but is required by
        \texttt{open\_pdks} and is conventionally bundled with the suite.
  \item \textbf{xschem} --- schematic capture
        (\url{https://github.com/StefanSchippers/xschem}). Also a separate
        upstream, included because \texttt{setup.sh} configures it.
  \item \textbf{open\_pdks} --- installer/wrapper for open-source PDKs such
        as \texttt{sky130A} and \texttt{gf180mcu}
        (\url{https://github.com/RTimothyEdwards/open_pdks}).
\end{itemize}

Everything will be installed under a single prefix:

\begin{lstlisting}
$OPEN_CIRCUITDESIGN_ROOT
|-- bin/        # magic, netgen, irsim, ngspice, xschem executables
|-- lib/        # shared libraries
|-- share/      # PDKs, scripts, technology files, magic/netgen/xschem libs
|-- include/    # development headers
\end{lstlisting}

\section{Prerequisites}

\subsection{Build dependencies (Debian / Ubuntu)}

Install the toolchain and X/Tcl libraries before building any of the tools.

\begin{lstlisting}
sudo apt-get update
sudo apt-get install -y \
    build-essential m4 tcsh csh git \
    tcl-dev tk-dev \
    libcairo2-dev \
    mesa-common-dev libglu1-mesa-dev \
    libncurses-dev \
    libx11-dev libxpm-dev libxt-dev \
    libreadline-dev \
    bison flex \
    autoconf automake libtool \
    libfftw3-dev libxaw7-dev \
    libxrender-dev libxcb-render0-dev libxcb-render-util0-dev \
    gawk \
    python3 python3-pip
\end{lstlisting}

\noindent The last three lines beyond the original toolchain bring in
xschem-specific dependencies (\texttt{libxrender}, the XCB render libs, and
\texttt{gawk}) along with Python, which is used by \texttt{open\_pdks}.

\subsection{Build dependencies (Red Hat / Fedora)}

\begin{lstlisting}
sudo dnf install -y \
    gcc gcc-c++ make m4 tcsh git \
    tcl-devel tk-devel \
    cairo-devel \
    mesa-libGL-devel mesa-libGLU-devel \
    ncurses-devel \
    libX11-devel libXpm-devel libXt-devel libXaw-devel \
    readline-devel \
    bison flex \
    autoconf automake libtool \
    fftw-devel \
    libXrender-devel \
    gawk \
    python3 python3-pip
\end{lstlisting}

\subsection{Create the install prefix and source directories}

Pick the install root once and use it throughout. The default in
\texttt{setup.sh} is \texttt{\$HOME/opencircuitdesign}; set it in your shell
\emph{first} so the build commands below pick it up:

\begin{lstlisting}
# Match the value in setup.sh (edit setup.sh first if you want a different path)
export OPEN_CIRCUITDESIGN_ROOT="$HOME/opencircuitdesign"

mkdir -p "$OPEN_CIRCUITDESIGN_ROOT"   # source trees clone in here directly
cd "$OPEN_CIRCUITDESIGN_ROOT"
\end{lstlisting}

\subsection{Use the included \texttt{setup.sh}}

Rather than hand-rolling \texttt{PATH}, \texttt{LD\_LIBRARY\_PATH},
\texttt{PDK\_ROOT}, and the per-tool environment variables, the repository
ships a \texttt{setup.sh} script that handles all of it. After confirming
the \texttt{OPEN\_CIRCUITDESIGN\_ROOT} line at the top of \texttt{setup.sh}
matches the prefix you chose above, source the script in any shell where
you want the tools available:

\begin{lstlisting}
# One-off: source it in the current shell
source /path/to/repo/setup.sh

# Persistent: have every login shell source it automatically
echo 'source /path/to/repo/setup.sh' >> ~/.bashrc
\end{lstlisting}

\noindent The script:

\begin{itemize}[leftmargin=2em]
  \item Verifies \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT} exists and bails out
        loudly if it doesn't.
  \item Prepends \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/bin} to \texttt{PATH}
        and \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/lib} to
        \texttt{LD\_LIBRARY\_PATH}.
  \item Sets \texttt{PDK\_ROOT} to
        \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/share/pdk} (the standard
        \texttt{open\_pdks} install location when configured with
        \texttt{-{}-with-pdk\_root=\$OPEN\_CIRCUITDESIGN\_ROOT/share/pdk}).
  \item Exports per-tool variables: \texttt{MAGIC\_HOME}, \texttt{NETGEN\_HOME},
        \texttt{NGSPICE\_SHARE\_DIR}, \texttt{XSCHEM\_SYSTEM\_LIBRARY\_PATH}.
  \item If \texttt{sky130A} is found under \texttt{\$PDK\_ROOT}, it also
        sets \texttt{PDK=sky130A}, \texttt{SKY130A},
        \texttt{MAGIC\_MAGICRC}, \texttt{NETGEN\_SETUP},
        \texttt{XSCHEM\_LIBRARY\_PATH}, and \texttt{XSCHEM\_RC}.
  \item Prints a short summary so you can confirm everything resolved
        correctly.
\end{itemize}

You should source \texttt{setup.sh} \emph{before} running the verification
commands at the end of each section below. The build steps themselves do
not need it --- they only need \texttt{OPEN\_CIRCUITDESIGN\_ROOT} exported.

\section{Magic}

Magic is the VLSI layout editor used as the geometry engine for the entire flow.

\begin{lstlisting}
cd "$OPEN_CIRCUITDESIGN_ROOT"
git clone https://github.com/RTimothyEdwards/magic.git
cd magic
./configure --prefix="$OPEN_CIRCUITDESIGN_ROOT"
make -j$(nproc)
make install
\end{lstlisting}

Verify (after \texttt{source setup.sh}):

\begin{lstlisting}
which magic
magic -dnull -noconsole -T scmos <<< "version; quit -noprompt"
\end{lstlisting}

\section{Netgen}

Netgen is the LVS (Layout-Versus-Schematic) and netlist comparison tool.

\begin{lstlisting}
cd "$OPEN_CIRCUITDESIGN_ROOT"
git clone https://github.com/RTimothyEdwards/netgen.git
cd netgen
./configure --prefix="$OPEN_CIRCUITDESIGN_ROOT"
make -j$(nproc)
make install
\end{lstlisting}

Verify:

\begin{lstlisting}
which netgen
netgen -batch <<< "puts [version]; quit"
\end{lstlisting}

\section{IRSIM}

IRSIM is a switch-level simulator that integrates with Magic.

\begin{lstlisting}
cd "$OPEN_CIRCUITDESIGN_ROOT"
git clone https://github.com/RTimothyEdwards/irsim.git
cd irsim
./configure --prefix="$OPEN_CIRCUITDESIGN_ROOT" \
            --with-tcl=/usr/lib/tcl8.6 \
            --with-tk=/usr/lib/tk8.6
make -j$(nproc)
make install
\end{lstlisting}

If your distribution uses a different Tcl/Tk version, adjust the
\texttt{-{}-with-tcl}/\texttt{-{}-with-tk} flags accordingly (or omit them to
let \texttt{configure} auto-detect).

Verify:

\begin{lstlisting}
which irsim
irsim -V
\end{lstlisting}

\section{ngspice}

ngspice is required by \texttt{open\_pdks} for simulation models and
verification. We build it from the official GitHub mirror at
\url{https://github.com/ngspice/ngspice} rather than the SourceForge
tarball, because SourceForge's file layout occasionally shifts and
hard-coded version URLs go stale; cloning sidesteps that entirely.

\begin{lstlisting}
cd "$OPEN_CIRCUITDESIGN_ROOT"
git clone https://github.com/ngspice/ngspice.git
cd ngspice
./autogen.sh
mkdir -p release && cd release
../configure --prefix="$OPEN_CIRCUITDESIGN_ROOT" \
             --enable-xspice \
             --enable-cider \
             --enable-openmp \
             --with-readline=yes \
             --disable-debug
make -j$(nproc)
make install
\end{lstlisting}

\noindent\textbf{Notes:}
\begin{itemize}[leftmargin=2em]
  \item \texttt{./autogen.sh} regenerates the \texttt{configure} script
        from \texttt{configure.ac}. The git tree does not ship a
        pre-generated \texttt{configure} the way release tarballs do, so
        this step is mandatory. It needs \texttt{autoconf},
        \texttt{automake}, and \texttt{libtool} (already in the
        prerequisite list).
  \item To build a tagged release rather than master, run
        \texttt{git tag -l} after cloning to list available tags, then
        \texttt{git checkout <tag>} (e.g.\ \texttt{git checkout ngspice-44})
        before running \texttt{autogen.sh}.
  \item \texttt{-{}-enable-xspice} is required for code-model support used by
        many open-source PDKs.
  \item Building inside a \texttt{release/} subdirectory keeps the source
        tree clean as recommended by the ngspice manual.
\end{itemize}

Verify:

\begin{lstlisting}
which ngspice
ngspice -v
\end{lstlisting}

\section{xschem}

xschem is the schematic capture tool used to drive ngspice simulations and
generate netlists for LVS against Magic layouts. It is a separate upstream
project from Tim Edwards' tools, but \texttt{setup.sh} configures the
xschem-related environment variables
(\texttt{XSCHEM\_SYSTEM\_LIBRARY\_PATH}, \texttt{XSCHEM\_LIBRARY\_PATH},
\texttt{XSCHEM\_RC}), so we install it into the same prefix.

\subsection{Verify xschem build dependencies}

If you followed the prerequisite lists in Section~2, you already have
everything xschem needs. The xschem-specific packages are:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Debian / Ubuntu:} \texttt{libxaw7-dev}, \texttt{libxrender-dev},
        \texttt{libxcb-render0-dev}, \texttt{libxcb-render-util0-dev},
        \texttt{tcl-dev}, \texttt{tk-dev}, \texttt{flex}, \texttt{bison},
        \texttt{libcairo2-dev}, \texttt{gawk}.
  \item \textbf{Red Hat / Fedora:} \texttt{libXaw-devel},
        \texttt{libXrender-devel}, \texttt{tcl-devel}, \texttt{tk-devel},
        \texttt{flex}, \texttt{bison}, \texttt{cairo-devel}, \texttt{gawk}.
\end{itemize}

\subsection{Clone, configure, build, and install}

\begin{lstlisting}
cd "$OPEN_CIRCUITDESIGN_ROOT"
git clone https://github.com/StefanSchippers/xschem.git
cd xschem
./configure --prefix="$OPEN_CIRCUITDESIGN_ROOT"
make -j$(nproc)
make install
\end{lstlisting}

\noindent\textbf{Notes:}
\begin{itemize}[leftmargin=2em]
  \item xschem ships a hand-written \texttt{configure} script (not autotools
        generated). \texttt{-{}-prefix} works as expected; most other
        autotools flags do not apply.
  \item If you need translucent rectangle / fill rendering, ensure the
        XCB render packages (\texttt{libxcb-render0-dev},
        \texttt{libxcb-render-util0-dev}) were installed before
        \texttt{configure} ran --- otherwise xschem falls back to a
        non-translucent code path.
  \item The build installs system symbol libraries to
        \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/share/xschem/}, which is what
        \texttt{setup.sh} points \texttt{XSCHEM\_SYSTEM\_LIBRARY\_PATH} at.
\end{itemize}

\subsection{Verify}

\begin{lstlisting}
which xschem
xschem --version

# Launch with no schematic to confirm the GUI starts:
xschem -b
\end{lstlisting}

\noindent After \texttt{open\_pdks} is installed (next section),
\texttt{setup.sh} will additionally export \texttt{XSCHEM\_LIBRARY\_PATH}
and \texttt{XSCHEM\_RC} pointing at the sky130A xschem libraries. To open
xschem with the sky130A symbol set, run:

\begin{lstlisting}
xschem --rcfile "$XSCHEM_RC"
\end{lstlisting}

\section{open\_pdks (install last)}

\texttt{open\_pdks} (\url{https://github.com/RTimothyEdwards/open_pdks}) is
a meta-installer: it fetches and configures vendor PDKs (sky130A, gf180mcu,
etc.) and stitches them into Magic, Netgen, ngspice, and xschem. \textbf{It
must be the last thing you build.} During \texttt{configure} and
\texttt{make install}, \texttt{open\_pdks} probes \texttt{PATH} for each
tool and runs them to generate tech files, magic \texttt{.magicrc} files,
netgen LVS setup files, and xschem libraries. If any of the tools is
missing or points at a different prefix, the resulting PDK install will be
incomplete or inconsistent.

\subsection*{Prerequisites for this section}

Before running anything in this section, confirm \emph{all} of the
following:

\begin{enumerate}[leftmargin=2em]
  \item Magic, Netgen, IRSIM, ngspice, and xschem are all installed under
        \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT} (Sections 3--7 of this guide).
  \item You have sourced \texttt{setup.sh} in the current shell:
        \begin{lstlisting}
source /path/to/repo/setup.sh
        \end{lstlisting}
  \item All four tools resolve under
        \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/bin}:
        \begin{lstlisting}
which magic netgen ngspice xschem
        \end{lstlisting}
        Each path printed should start with the value of
        \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT}. If any of them is missing or
        points at \texttt{/usr/bin}, fix that \emph{first} --- do not
        proceed with \texttt{open\_pdks}.
\end{enumerate}

\subsection{Configure and build}

With just \texttt{-{}-prefix}, \texttt{open\_pdks} installs the PDKs under
\texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/share/pdk}, which is exactly where
\texttt{setup.sh} points \texttt{PDK\_ROOT}. Magic, Netgen, and ngspice
are auto-detected from \texttt{PATH}, so the explicit
\texttt{-{}-with-magic=}, \texttt{-{}-with-netgen=}, and
\texttt{-{}-with-ngspice=} flags are unnecessary as long as you sourced
\texttt{setup.sh} (or otherwise put \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/bin}
on \texttt{PATH}) before running \texttt{configure}.

\textbf{Important:} the sky130 and gf180mcu PDKs must be configured and
installed \emph{separately} --- run \texttt{configure}, \texttt{make}, and
\texttt{make install} for one PDK before moving on to the other. Do not
combine both \texttt{-{}-enable-*-pdk} flags in a single \texttt{configure}
invocation.

Start by cloning \texttt{open\_pdks}:

\begin{lstlisting}
cd "$OPEN_CIRCUITDESIGN_ROOT"
git clone https://github.com/RTimothyEdwards/open_pdks.git
cd open_pdks
\end{lstlisting}

\subsubsection*{sky130 PDK}

The sky130 \texttt{configure} pulls in the base sky130 PDK along with the
OpenRAM SRAM macros (\texttt{-{}-enable-sram-sky130}), the SRAM space
variant (\texttt{-{}-enable-sram-space-sky130}), the ReRAM macros
(\texttt{-{}-enable-reram-sky130}), and the OSU standard-cell libraries at
the 12T, 15T, and 18T track heights
(\texttt{-{}-enable-osu-t12-sky130}, \texttt{-{}-enable-osu-t15-sky130},
\texttt{-{}-enable-osu-t18-sky130}).

\begin{lstlisting}
./configure --prefix="$OPEN_CIRCUITDESIGN_ROOT" \
            --enable-sky130-pdk \
            --enable-sram-sky130 \
            --enable-osu-t12-sky130 \
            --enable-osu-t15-sky130 \
            --enable-osu-t18-sky130 \
            --enable-sram-space-sky130 \
            --enable-reram-sky130
make
make install
\end{lstlisting}

\subsubsection*{gf180mcu PDK}

Once the sky130 install completes, configure and build the gf180mcu PDK.
This pulls in the base gf180mcu PDK along with the Avalon standard-cell
library (\texttt{-{}-enable-avalon-sc-gf180mcu}), the OCD I/O library
(\texttt{-{}-enable-ocd-io-gf180mcu}), and the OCD SRAM macros
(\texttt{-{}-enable-ocd-sram-gf180mcu}).

\begin{lstlisting}
./configure --prefix="$OPEN_CIRCUITDESIGN_ROOT" \
            --enable-gf180mcu-pdk \
            --enable-avalon-sc-gf180mcu \
            --enable-ocd-io-gf180mcu \
            --enable-ocd-sram-gf180mcu
make
make install
\end{lstlisting}

The \texttt{--enable-*-pdk} flags select which PDK to fetch on each pass.
Drop any sub-options you don't need; each PDK is several gigabytes after
install.

\subsection{Install notes (each PDK takes 30--90 minutes)}

The PDKs land under \texttt{\$PDK\_ROOT}, e.g.\
\texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/share/pdk/sky130A} and
\texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/share/pdk/gf180mcuD}. Once
\texttt{sky130A} exists, sourcing \texttt{setup.sh} again will pick it up
and export \texttt{PDK}, \texttt{SKY130A}, \texttt{MAGIC\_MAGICRC},
\texttt{NETGEN\_SETUP}, \texttt{XSCHEM\_LIBRARY\_PATH}, and
\texttt{XSCHEM\_RC} automatically.

\subsection{Verify}

\begin{lstlisting}
source /path/to/repo/setup.sh

ls "$PDK_ROOT"
# Expect: sky130A  sky130B  gf180mcuA  gf180mcuB  gf180mcuD  gf180mcuD  ...

# setup.sh should now report sky130A as the selected PDK:
echo "$PDK"           # -> sky130A
echo "$MAGIC_MAGICRC" # -> .../sky130A/libs.tech/magic/sky130A.magicrc
echo "$NETGEN_SETUP"  # -> .../sky130A/libs.tech/netgen/sky130A_setup.tcl

# Open a sky130 layout in magic to confirm tech files were installed:
magic -rcfile "$MAGIC_MAGICRC"
\end{lstlisting}

\section{Tutorial repository}

A companion tutorial for using this tool stack to design analog/digital
circuits in the Global Foundries 180\,nm MCU process is available at:

\begin{center}
\url{https://github.com/stineje/gf180mcu-open-source-vlsi}
\end{center}

\noindent The repository ships with the \texttt{setup.sh} this guide
references, schematic and layout examples, simulation testbenches, and
step-by-step instructions for the canonical Magic / Netgen / xschem /
ngspice flow against the gf180mcu PDK.

\subsection{Clone the tutorial}

Pick any directory you like for tutorial work --- it does \emph{not} need
to live under \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT}. A common choice is a
class or coursework folder:

\begin{lstlisting}
cd ~                  # or wherever you keep coursework
git clone https://github.com/stineje/gf180mcu-open-source-vlsi.git
cd gf180mcu-open-source-vlsi
git checkout ieee_chipathon
\end{lstlisting}

\noindent The \texttt{ieee\_chipathon} branch holds the IEEE ChipAThon
materials for this course. The default branch (\texttt{main}) tracks
upstream development; for coursework, stay on
\texttt{ieee\_chipathon} for now..

\subsection{Source the tutorial's \texttt{setup.sh}}

The tutorial's \texttt{instructions-setup/setup.sh} is the same script
described in this guide. Source it from your shell so all the tool and
PDK environment variables are set:

\begin{lstlisting}
source ./instructions-setup/setup.sh
\end{lstlisting}

\noindent To make this automatic for every login shell, append the line
to your \texttt{\textasciitilde/.bashrc} with the absolute path:

\begin{lstlisting}
echo "source $HOME/gf180mcu-open-source-vlsi/instructions-setup/setup.sh" \
     >> ~/.bashrc
\end{lstlisting}

\subsection{Verify the tutorial environment}

After sourcing, the script prints a summary. Confirm that
\texttt{PDK\,=\,gf180mcuD} (or whatever variant you have installed) and
that \texttt{magic}, \texttt{netgen}, \texttt{ngspice}, and
\texttt{xschem} all resolve under \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/bin}.
If \texttt{PDK} reports \texttt{sky130A} instead of a gf180mcu variant,
your gf180mcu install didn't complete --- revisit Section~8 before
working through the tutorial.

\subsection{Keeping up to date}

The tutorial is updated periodically with new examples and corrections.
To pull the latest on the \texttt{ieee\_chipathon} branch:

\begin{lstlisting}
cd ~/gf180mcu-open-source-vlsi
git checkout ieee_chipathon    # confirm you're on the course branch
git pull
\end{lstlisting}

\section{Putting it all together}

\subsection{Quick sanity check}

After every tool is installed and you have sourced \texttt{setup.sh},
the script's own summary output should list each tool's location. You can
also run:

\begin{lstlisting}
for tool in magic netgen irsim ngspice xschem; do
    printf "%-8s -> %s\n" "$tool" "$(which $tool 2>/dev/null || echo '<not found>')"
done
\end{lstlisting}

All paths should resolve under \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/bin}.

\subsection{Recommended directory layout after install}

\begin{lstlisting}
$OPEN_CIRCUITDESIGN_ROOT
|-- bin/
|   |-- magic
|   |-- netgen
|   |-- irsim
|   |-- ngspice
|   `-- xschem
|-- include/
|-- lib/
|-- share/
|   |-- magic/      # MAGIC_HOME
|   |-- netgen/     # NETGEN_HOME
|   |-- ngspice/    # NGSPICE_SHARE_DIR
|   |-- xschem/     # XSCHEM_SYSTEM_LIBRARY_PATH
|   |-- pdk/        # PDK_ROOT
|   |   |-- sky130A/
|   |   `-- gf180mcuA/
|   `-- man/
|-- magic/          # cloned source trees (sit alongside bin/, lib/, share/)
|-- netgen/
|-- irsim/
|-- ngspice/
|-- xschem/
`-- open_pdks/
\end{lstlisting}

\section{Troubleshooting}
\label{sec:troubleshooting}

\begin{description}[leftmargin=0em, style=nextline]
  \item[\texttt{ERROR: Directory \$OPEN\_CIRCUITDESIGN\_ROOT does not exist}]
    \texttt{setup.sh} aborts when its target prefix is missing. Either
    create the directory (\texttt{mkdir -p "\$OPEN\_CIRCUITDESIGN\_ROOT"})
    or edit the \texttt{OPEN\_CIRCUITDESIGN\_ROOT=...} line at the top of
    \texttt{setup.sh} to point somewhere that does exist.

  \item[\texttt{configure: error: Tcl headers not found}]
    Make sure \texttt{tcl-dev} / \texttt{tcl-devel} is installed and that
    \texttt{tclConfig.sh} lives in a standard location (typically
    \texttt{/usr/lib/tclX.Y/}). Pass \texttt{-{}-with-tcl=<dir>} explicitly
    if needed.

  \item[Magic launches but graphics are missing]
    Build Magic with Cairo and OpenGL support enabled (the default when
    the corresponding \texttt{-dev} packages are present). Re-run
    \texttt{./configure} after installing them and rebuild.

  \item[\texttt{open\_pdks} fails to find Magic / Netgen]
    Ensure \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/bin} is on \texttt{PATH}
    before running \texttt{./configure} (sourcing \texttt{setup.sh} does
    this). If auto-detection still fails, pass absolute paths explicitly
    via \texttt{-{}-with-magic=\$OPEN\_CIRCUITDESIGN\_ROOT/bin/magic},
    \texttt{-{}-with-netgen=\$OPEN\_CIRCUITDESIGN\_ROOT/bin/netgen}, and
    \texttt{-{}-with-ngspice=\$OPEN\_CIRCUITDESIGN\_ROOT/bin/ngspice}.

  \item[\texttt{setup.sh} warns ``\texttt{No PDK directory found}'']
    \texttt{open\_pdks} hasn't been built/installed yet, or it was
    configured with a \texttt{-{}-prefix} that doesn't match
    \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT}. Re-run \texttt{./configure} for
    \texttt{open\_pdks} with
    \texttt{-{}-prefix="\$OPEN\_CIRCUITDESIGN\_ROOT"}, or override the PDK
    location explicitly with
    \texttt{-{}-with-pdk\_root="\$OPEN\_CIRCUITDESIGN\_ROOT/share/pdk"}.

  \item[Shared libraries not found at runtime]
    Source \texttt{setup.sh} (it sets \texttt{LD\_LIBRARY\_PATH}). For a
    system-wide alternative, add \texttt{\$OPEN\_CIRCUITDESIGN\_ROOT/lib}
    to \texttt{/etc/ld.so.conf.d/opencircuitdesign.conf} and run
    \texttt{sudo ldconfig}.

  \item[Updating later]
    Each tool can be rebuilt independently:
    \texttt{cd \$OPEN\_CIRCUITDESIGN\_ROOT/<tool> \&\& git pull \&\& make \&\& make install}.
    Re-run \texttt{open\_pdks} only when changing PDK selection or after a
    significant Magic/Netgen update.
\end{description}

\section{References}

\begin{itemize}[leftmargin=2em]
  \item Open Circuit Design home page: \url{http://opencircuitdesign.com}
  \item Magic on GitHub: \url{https://github.com/RTimothyEdwards/magic}
  \item Netgen on GitHub: \url{https://github.com/RTimothyEdwards/netgen}
  \item IRSIM on GitHub: \url{https://github.com/RTimothyEdwards/irsim}
  \item ngspice: \url{https://ngspice.sourceforge.io}
  \item xschem on GitHub: \url{https://github.com/StefanSchippers/xschem}
  \item open\_pdks on GitHub: \url{https://github.com/RTimothyEdwards/open_pdks}
  \item gf180mcu open-source VLSI tutorial: \url{https://github.com/stineje/gf180mcu-open-source-vlsi}
  \item IIC-OSIC-TOOLS container: \url{https://github.com/iic-jku/iic-osic-tools}
\end{itemize}

\appendix

\section{Container alternative for macOS and Windows}
\label{app:container}

The build-from-source instructions in Sections~3--8 target Linux. On
macOS and Windows, native builds of Magic, xschem, and the rest of the
stack are painful (or impossible without WSL). The recommended
alternative is the \textbf{IIC-OSIC-TOOLS} Docker container from the
Institute for Integrated Circuits at TU Wien, which bundles Magic,
Netgen, IRSIM, xschem, ngspice, KLayout, OpenROAD, yosys, and a
pre-installed PDK set (sky130A, gf180mcuD, etc.) into a single image.
Repository: \url{https://github.com/iic-jku/iic-osic-tools}.

This appendix covers running the container on macOS and Windows. Linux
users can also use it, but the from-source path documented earlier is
usually preferable on Linux.

\subsection{Prerequisites}

\textbf{macOS:} Install Docker Desktop from
\url{https://www.docker.com/products/docker-desktop/}. Launch it once
after install so the engine starts; you should see ``Engine running'' in
the lower-left of the Docker Desktop window.

\textbf{Windows:} Install Docker Desktop with the WSL2 backend (the
default on Windows~10 and~11). If WSL2 isn't already installed, open
PowerShell as administrator and run:

\begin{lstlisting}
wsl --install
\end{lstlisting}

\noindent then reboot. After install, launch Docker Desktop once.

\textbf{Both platforms:} you will run shell commands. On macOS, open the
\textbf{Terminal} app (Spotlight: \texttt{Cmd+Space}, type
\texttt{Terminal}). On Windows, open the WSL Ubuntu terminal (Start
menu) or use Docker Desktop's built-in shell (the \texttt{>\_} icon at
the bottom of the Docker Desktop window). Verify Docker is reachable:

\begin{lstlisting}
docker --version
\end{lstlisting}

\subsection{Apple Silicon caveat}

The \texttt{hpretl/iic-osic-tools} image is built for x86\_64. On
Apple Silicon (M1/M2/M3/M4), Docker Desktop runs it under emulation:
roughly 2--3$\times$ slower than native and uses more RAM, but
functional. You may see a warning the first time the image is pulled:

\begin{lstlisting}
WARNING: The requested image's platform (linux/amd64) does not match
         the detected host platform (linux/arm64/v8)
\end{lstlisting}

\noindent That is expected; you can pin the platform explicitly to
suppress it:

\begin{lstlisting}
docker pull --platform linux/amd64 hpretl/iic-osic-tools:latest
\end{lstlisting}

Intel Macs and Windows hosts have no equivalent issue.

\subsection{Clone the wrapper scripts and start a VNC session}

The container's GUI tools are exposed via a browser-based VNC desktop,
which works identically on macOS and Windows and avoids the X11
forwarding problems of XQuartz / VcXsrv. From your terminal:

\begin{lstlisting}
cd ~
git clone https://github.com/iic-jku/iic-osic-tools.git
cd iic-osic-tools
./start_vnc.sh
\end{lstlisting}

The first run pulls the image (roughly 10\,GB --- expect 5 to 45 minutes
depending on bandwidth). You can watch progress under
\textbf{Images} in Docker Desktop's left sidebar. When the script
finishes starting the container, it prints a URL similar to:

\begin{lstlisting}
http://localhost:8888/vnc.html?autoconnect=true&resize=remote
\end{lstlisting}

Open that URL in any browser. You will land on a full XFCE desktop with
Magic, xschem, KLayout, ngspice, Netgen, and IRSIM all installed. PDKs
are at \texttt{/foss/pdks/} inside the container ---
\texttt{sky130A}, \texttt{gf180mcuD}, etc.

\subsection{Mounting your work into the container}

By default the container's \texttt{/foss/designs} directory is bound to
\texttt{\textasciitilde/iic-osic-tools-volume/} on the host. Anything
saved there from inside the container persists on your machine.

To mount a different host folder --- say a clone of the gf180mcu
tutorial --- pass the path as an argument:

\begin{lstlisting}
./start_vnc.sh ~/work/gf180mcu-open-source-vlsi
\end{lstlisting}

\noindent On Windows from WSL2, refer to Windows-side paths via the
\texttt{/mnt/c/} mount:

\begin{lstlisting}
./start_vnc.sh /mnt/c/Users/jstine/Documents/work
\end{lstlisting}

\subsection{Using the gf180mcu tutorial inside the container}

The PDK layout inside the container matches what \texttt{setup.sh}
expects, but the container is already pre-configured --- you do not need
to source \texttt{setup.sh}. Inside the VNC desktop, open a terminal and
run:

\begin{lstlisting}
cd /foss/designs
git clone https://github.com/stineje/gf180mcu-open-source-vlsi.git
cd gf180mcu-open-source-vlsi
git checkout ieee_chipathon

export PDK_ROOT=/foss/pdks
export PDK=gf180mcuD
magic -rcfile $PDK_ROOT/$PDK/libs.tech/magic/$PDK.magicrc
\end{lstlisting}

\subsection{Improving the VNC experience}

The browser-based VNC is fine for one or two windows. For serious
schematic editing, the browser sometimes intercepts keyboard shortcuts.
Two fixes:

\begin{itemize}[leftmargin=2em]
  \item Press \texttt{F11} in the browser to enter fullscreen mode --- this
        frees most shortcuts.
  \item Connect with a native VNC client instead. The
        \texttt{start\_vnc.sh} script also exposes a raw VNC port
        (typically \texttt{localhost:5901}). On macOS, the built-in
        \textbf{Screen Sharing} app handles \texttt{vnc://localhost:5901}
        cleanly; on Windows, install TigerVNC or RealVNC.
\end{itemize}

\subsection{Stopping, restarting, and updating}

The container keeps running until you stop it. From the host terminal:

\begin{lstlisting}
docker ps                       # list running containers
docker stop iic-osic-tools      # stop the container
\end{lstlisting}

\noindent or click the stop button next to the container in the Docker
Desktop \textbf{Containers} tab. To use it again later:

\begin{lstlisting}
cd ~/iic-osic-tools && ./start_vnc.sh
\end{lstlisting}

\noindent To pull a newer release of the image (the project ships
periodic updates):

\begin{lstlisting}
docker pull hpretl/iic-osic-tools:latest
./start_vnc.sh
\end{lstlisting}

\noindent Files in your mounted work directory are preserved across
image updates; only the container's read-only OS layer changes.

\subsection{When to use this versus the from-source build}

The container gets you running in minutes and isolates the entire stack
from the host. Trade-offs to keep in mind:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Pros:} no native build dependencies; identical environment
        across macOS / Windows / Linux teammates; one command to update;
        easy to throw away and recreate.
  \item \textbf{Cons:} ~10\,GB disk; GUI feels less snappy over VNC than a
        native window; tool versions follow the container release cadence
        (you cannot trivially patch Magic and rebuild without forking the
        image); on Apple Silicon, x86\_64 emulation has a measurable
        performance cost.
\end{itemize}

For class assignments and tutorial work on macOS or Windows, the
container is almost always the right choice. Building from source
(Sections~3--8) is preferable when you are hacking on the tools
themselves, debugging a specific build issue (such as the gf180mcu
\texttt{(UNNAMED)} annotation error discussed in
Section~\ref{sec:troubleshooting} \emph{Troubleshooting}), or working on
Linux with full root access to the host.

\end{document}
